### 2023-05-05 misc/expo/additive.tex
Bogolyubov

\def\FF{{\bf F}}
\def\TT{{\bf T}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}

@@

\maketitle{Notes on additive combinatorics}{by}{Marcel K. Goh}{23 March 2022}

@@

>>>>

\def\FF{{\bf F}}
\def\TT{{\bf T}}
\def\bar{\overline}
\def\hat{\widehat}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\mone{-\!\one}  % aesthetic quibble

@@

\maketitle{Notes on additive combinatorics}{by}{Marcel K. Goh}{5 May 2023}

@@

\advsect Bohr sets and Bogolyubov's lemma

Let $G$ be a finite abelian group, let $\chi_1, \ldots, \chi_k$ be characters on $G$, and let
$\delta>0$. The {\it Bohr set} $B(\chi_1,\ldots,\chi_k;\delta)$ is the set
$$\bigl\{ x\in G : \chi_i(x)\in e\bigl([-\delta,\delta]\bigr)\ \hbox{for all}\ 1\le i\le k\bigr\}.$$
(As before, we write $e(\alpha)$ for $e^{2\pi i\alpha}$.) We say that such a Bohr set has {\it dimension} $k$
and {\it width} $\delta$.

If $G= \FF_p^n$, we can write $\chi_i = \chi_{u_i}$ where $\chi_{u_i}(x) = e(u_i\cdot x/p$. If $\delta<1/p$,
then $\chi_{u_i}(x)\in e\bigl([-\delta,\delta]\bigr)$ if and only if
$(u_i\cdot x) / p \bmod 1\in [-\delta,\delta]$. But $u_i\cdot x$ is an integer, so this is true if and only
if $u_i\cdot x = 0$, so consequently
$$B(\chi_1,\ldots,\chi_k;\delta) = \{ x\in \FF_p^n : u_i\cdot x = 0\ \hbox{for all}\ 1\le i\le k\},$$
which is a subspace cut out by $k$ linear conditions, i.e., of codimension at most $k$.

\parenproclaim Lemma {\advthm} (Bogolyubov). Let $G$ be a finite abelian group. Let $A\subseteq G$ be a
subset of density $\alpha$. Then $2A-2A$ contains a Bohr set of dimension at most $1/\alpha^2$ and width
$1/4$.

\proof For $x\in G$, let
$$\eqalign{
f(x) &= \bigl( \one_A * \one_A * (\mone_A) * (\mone_A)\bigr)(x) \cr
&= \ex_{p+q-r-s=x} \one_A(p)\one_A(q)\one_A(r)\one_A(s).
}$$
Then $x\in 2A-2A$ if and only if $f(x)\ne 0$. We want to find a Bohr set on which $f$ does not vanish.

Note first that for any character $\chi$,
$$\hat{\mone_A}(\chi) = \ex_x \one_A(-x)\bar{\chi(x)} = \ex_x \one_A(x)\bar{\chi(-x)} = \ex_x\one_A(x)\chi(x)
=\bar{\hat{\one_A}(\chi)}.$$
(The last identity relies on the fact that $\one_A(x)$ is real.) This implies that
$\hat{\one_A}(\hat{\mone_A}) = |\one_A|^2$. Now we apply the Fourier inversion
formula and the convolution law to expand
$$\eqalign{
f(x) &= \sum_\chi \hat{\one_A*\one_A*(\mone_A)*(\mone_A)}(\chi)(\chi)(x) \cr
&= \sum_\chi \hat{\one_A}(\chi)^2 \hat{\mone_A}(\chi)^2 \chi(x) \cr
&= \sum_\chi \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x). \cr
}$$
Let
$$K = \bigl\{ \chi\in \hat G : \bigl| \hat{\one_A}(\chi) \bigr| \ge \alpha^{3/2}\bigr\}$$
and let $B$ be the Bohr set $B(K;1/4)$. Then the expression above for $f(x)$ can be split into
$$f(x) =
\bigl| \hat{\one_A}(\chi_0) \bigr|^4
+ \sum_{\chi\in K} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)
+ \sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x).$$
We shall assume that $x\in B$ and deal with each of the three terms separately.

The first term is easy; since $A$ has density $\alpha$, $\hat{\one_A}(\chi_0) = \alpha$, and thus the first
time is $\alpha^4$. For the second term, note that if $\chi\in K$ and $x\in B$, then
$\chi(x)\in e\bigl([-1/4, 1/4]\bigr)$. This means that $\chi(x)$ has argument between $-\pi/2$ and $\pi/2$,
and {\it a fortiori} has nonnegative real part. Now, if $\chi\notin K$ and $\chi\ne \chi_0$,
then $\bigl|\hat{\one_A}\bigr| < \alpha^{3/2}$, so we take the absolute value of the third term
and apply the triangle inequality and Parseval's identity to obtain
$$\Bigl|\sum_{\chi\notin K\cup \{\chi_0\}} \bigl| \hat{\one_A}(\chi) \bigr|^4 \chi(x)\Bigr|
\le \sum_{\chi\notin K\cup\{\chi_0\}} \bigl|\hat{\one_A}(\chi)\bigr|^4
< \sum_{\chi\in \hat G} \alpha^3 \bigl| \hat{\one_A}(\chi)\bigr|^2
= \alpha^3\langle \hat{\one_A},\hat{\one_A}\rangle.$$
But by Parseval's identity, the right-hand side equals $\alpha^3 \langle \one_A,\one_A\rangle = \alpha^4$,
so we find that the real part of the third term is more than $-\alpha^4$. This means that $\Re f(x) > 0$,
so $f(x)\ne 0$ for all $x\in B$; that is, $B\subseteq 2A-2A$.

It remains to give an upper bound on the size of $K$. We have
$$\alpha = \sum_{\chi\in \hat G} \bigl| \one_A(\chi)\bigr|^2 \ge
\sum_{\chi\in K} \bigl|\hat{\one_A(\chi)}\bigr|^2 \ge |K|\alpha^3,$$
meaning that $|K|\le 1/\alpha^2$.\slug

When $A$ is a subset of density $\alpha$ in $G = \FF_p^n$, Bogolyubov's lemma tells us that $2A-2A$ contains
a subspace of codimension at most $1/\alpha^2$.
